TODO: Title

### 2013-02-21 Doc/Miccai2013/miccai2013.tex
more on the miccai paper
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@513 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@

The Glisson's capsule \cite{Umale2013} has in order of three higher
stiffness than the parenchyma. As such it plays an important role in the
deformation of the liver \cite{Ahn2010,Hollenstein2006}. Evaluation of the
liver as a homogeneous material can lead to overestimation of the
mechanical properties up to the factor of 3 \cite{Hollenstein2006}.

@@

% -- TODO: what is membrane element
% --

@@

@@

\end{document}
% vim:set et sw=2 tw=75 fdm=marker fdl=1 fdc=4 isk+=_,-:

>>>>

%%% Comments
\usepackage{color}
\newcommand{\TG}[1]{{\color{blue}\textbf{TG: #1}}}
%%%

\newcommand{\Vec}[1]{\mathbf{#1}}
\newcommand{\Mat}[1]{\mathbf{#1}}

@@

% -- TODO why model liver? what's interesting about it?
% -- liver covered by capsule, thin membrane

the usual approach is to model the organ as homogeneous material,
as such complex non-linear material descriptions are inevitably necessary
to describe the properties of non-homogeneous tissue.

modeling non-homogenous volumetric object by a dense mesh is complicated by
large number of elements. we propose a composite model and show that it
leads to adequate results.

The Glisson's capsule has in order of three higher
stiffness than the parenchyma. As such it plays an important role in the
deformation of the liver \cite{Ahn2010,Hollenstein2006}. Evaluation of the
liver as a homogeneous material can lead to overestimation of the
mechanical properties up to the factor of 3 \cite{Hollenstein2006}.

@@

GOAL: show that we are able to simulate the behaviour of non-homogeneous
tissue covered by capsule (liver/kidney) by combined model.

GOAL2: show that combination of linear tetrahedra and linear membrane is
easy.

% NOTE: the assumption is that we cannot estimate elastic modulus as a
% combination of the two moduli (for parenchyma and capsule) because it will
% work only for certain displacement. Is this valid?
% -- based on the text of Hollenstein2006 it seems like it

\section{Model}

\subsection{parenchyma} %{{{

\TG{TODO: write something like this} % (src: Peterlik2012)
\begin{quote}
While most studies concerning the material properties of liver parenchyma
agree on a viscoelastic behaviour (see [5] or [7] for instance), we employ
a simpler non-linear elastic model. This is essentially motivated by the
fact that we are not focusing on the transient part of the deformation but
rather the static equilibrium under some specific loading conditions.
\end{quote}

We model the parenchyma using corotational finite elmets. The corotational
formulation allows us to deal with geometrical non-linearities but still
rely on linear stress-strain relationship. Thus rotations and large
deformations are possible. In the corotational formulation the deformation
of each element is expressed in the local frame of reference. There are
several ways of computing the corotational frame for elements, we rely on
the geometrical method proposed in \cite{Nesme2005}.
% NOTE: This is very simplified and could be extended.
The stiffness matrix $\Mat{K}$ can be expressed in terms of the deformation
$u$ and the force as $\Vec{f} = \Mat{K}(\Vec{u}) \Vec{u}$. The resulting
system of linear equations is then solved by direct LDL solver.

%}}}

\subsection{Capsule} %{{{

The thickness of the Glisson's capsule is relatively small (Umale et al.
\cite{Umale2011} reported for porcine liver values in range of 10-20
$\mu$m). This makes it very difficult to model with volumetric mesh. Such
mesh would inevitably contain large number of elements or very irregular
elements. Instead modeling the capsule with two-dimensional elements seems
as a natural choice. Mechanics provides a FE tool in the form of membranes
and shells. Based on the small thickness (compared to the area) we also
assume negligible bending forces and propose a model based on membrane
elements.

To maintain simplicity of the composite model we choos simple triangular
elements with constant strain.

\TG{add a ref. or dive into the definition?}

We again use linear elastic model and employ the  corotational formulation
for the triangular elements.

%}}}

\subsection{Coupling Between Capsule and Parenchyma} %{{{

High cohesion between capsule and parenchyma was reported. Based on this
property we may assume no relative motion of the capsule to the parenchyma.

While in general we could use arbitrary surface mesh for a capsule we use
the fact that the parenchyma is modelled by tetrahedral elements which have
triangular faces. Because the boundary of the volumetric mesh is already
triangulated, we only extract the boundary triangles which we use for the
model of capsule.

Using directly the boundary of the tetrahedral mesh does not only solve the
problem of obtaining the triangular mesh, but has one more advantage. Nodes
of the triangular mesh overlap with nodes of the tetrahedral mesh and we do
not have to project nodes of tiangular mesh onto the tetrahedra. More over,
the stiffness matrices for capsule and parenchyma are easily assembled
together and solved as one system.

\TG{Is assembling process too obvious?}

Withouth the loss of generality we can assume the tetrahedron consists of
nodes $p_1, p_2, p_3$ and $p_4$, the boundary triangle has nodes $p_1, p_2$
and $p_3$. We can reorder the degrees of freedom so that the stiffness
matrix $\Mat{K}^t$ for the tetrahedron can be written as:

\begin{equation}
  \Mat{K}^t = \left[\begin{array}{c|c}
      \Mat{K}^t_{1-3,1-3} & \Mat{K}^t_{1-3,4} \\
      \hline
      \Mat{K}^t_{4,1-3} & \Mat{K}^t_{4,4} \\
  \end{array}\right]
\end{equation}

Then the assembled stiffness matrix for the element is:

\begin{equation}
  \Mat{K} = \left[\begin{array}{c|c}
      \Mat{K}^t_{1-3,1-3} & \Mat{K}^t_{1-3,4} \\
      \hline
      \Mat{K}^t_{4,1-3} & \Mat{K}^t_{4,4} \\
  \end{array}\right]
  +
  \left[\begin{array}{c|c}
      \Mat{K}^m & 0 \\
      \hline
      0 & 0 \\
  \end{array}\right]
\end{equation}

Where $\Mat{K}^m$ is the stiffness matrix of the triangular membrane.

%}}}

\subsection{Blood Vessels} %{{{

We also combined our composite model with vascularization method proposed
by Peterlik et al. \cite{Peterlik2012}. The method is based on linear beams
with local frames of reference (in a many aspects similar to the
corotational formulation). As such the model also handles geometric
non-linearities in the deformation. The proposed model assumes no relative
motion of vessels to the parenchyma. Segmented vascular network is
discretized into the set of linear elements that are then projected into
the tetrahedral mesh of the parenchyma. During the simulation forces and
torques resulting from the deformations of vessels by parenchyma are
computed and propagated back onto the tetrahedral mesh.

%}}}

@@

setup:

Test 1) Global deformation of liver under gravity
  * show that there is a difference between simple homogeneous material
    and material with capsule
  * include blood vessels

Test 2) Indentation test
  * perform indentation and analyze the profile
  * show how stiffer membrane affect the result

@@

\end{document}
% vim:set et sw=2 tw=75 fdm=marker fdl=3 fdc=4 isk+=_,-:
